Donor Responses to Civil Society Repression

### 2024-11-05 manuscript.qmd
Final Suparna edits

Beginning in 2003, Africa and Middle East Refugee Assistance (AMERA) International provided pro bono legal support for refugees entering Egypt. Based in London, this international non-governmental organization (INGO) maintained a small staff of legal experts in Cairo and raised the bulk of its funding from organizations in London, receiving the bulk of its revenue from a private charitable foundation, Comic Relief UK. Despite laws restricting INGO activities, AMERA enjoyed a positive relationship with government authorities. In the wake of the 2011 Arab Uprisings, though, the Egyptian government began strictly enforcing laws prohibiting foreign funding for civil society, blocking AMERA's access to Comic Relief donations. In 2014, AMERA shuttered its Egyptian office and redirected its programming to local projects in the United Kingdom [@Heiss:2019a].

The experiences of NGOs like AMERA are common. While a large literature in law and political science focuses on the role of international law, domestic legal institutions, and legal mobilization that may reduce repression [@Klug:2005; @PowellStaton:2009; @Simmons:2009; @ConradRitter:2013], the last two decades have shown that states also use laws not to expand rights, but to further restrict rights and repress organizations [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020; @GlasiusSchalkDeLange:2020; @BromleySchoferLonghofer:2020; @Chaudhry:2022]. Administrative crackdown—or the use of laws to create barriers to entry, funding, and advocacy by civil society groups—has proliferated across the globe over the past few decades—both across democracies and autocracies [@Chaudhry:2022]. As of 2023, 87% of the world's population now lives in countries where the space available to civil society is closed, repressed, or obstructed [@CIVICUSMonitor].

The growth in states using law to repress can be attributed to three main factors. First, administrative or legal crackdowns have many advantages over the use of violence. It allows governments to maintain a democratic facade abroad and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022]. Anti-NGO laws rarely elicit international condemnation or threats to withdraw aid. States also learn about the efficacy of these legal strategies from their peers and subsequently implement them in their own country [@GlasiusSchalkDeLange:2020; @Gilbert:2020; @Chaudhry:2022]. Finally, these tactics are part of a broader trend of using legal means to subvert democratic institutions and norms [@LevitskyZiblatt:2018].

Administrative crackdown via anti-NGO laws that bar access to foreign funding and foreign technical assistance has been particularly insidious. Without access to foreign funds, many INGOs cannot continue their overseas operations. Even domestic NGOs may be heavily strapped for money. Most NGOs in the Global South require a constant stream of competitive grants from foreign donors to be able to implement specific projects [@Bush:2015]. Local groups tackling contentious issues may not be able to raise funds domestically, as publics may prefer donating to service organizations focusing on health, education, and poverty [@DupuyRonPrakash:2016; @Brechenmacher:2017]. Even local philanthropists may be deterred from giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution.

@@

# State responses

A rich literature on the purposes of foreign aid explores why states transfer money and resources to others. Bilateral national aid agencies and multilateral international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with. States rhetorically declare normative justifications for engaging in development, with recipient-focused development goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011]. In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy-oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017]. A key enforcement mechanism to ensure that recipient-country actions align with donor-country goals is to condition aid on the achievement of specific democratizing and economic goals [@CheesemanSwedlundOBrien-Udry:2024; @HoefflerJustino:2024]. However, this dual-purpose strategy often creates dilemmas—if aid is withdrawn, it may jeopardize humanitarian programs or worsen political conditions. Conversely, continuing aid might signal tolerance for rights violations, weakening the credibility of the donor's human rights agenda. Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that donor agencies punish repressive states by reducing or withdrawing aid [@LebovicVoeten:2009], increasing aid to repressive states as a way to counter state violence [@Nielsen:2013], or making no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b].

One reason for such mixed results is that the decision to withdraw or suspend aid in response to human rights abuses is often seen as a binary choice driven by self-interested motivations: a donor country that sees human rights abuses in a recipient state will consider its own foreign policy interests before deciding to either continue or curtail aid. In reality, however, the decision is far more complex and involves a wide range of possible motivations and responses. @CheesemanSwedlundOBrien-Udry:2024 offer a stylized decision tree of possible choices following recipient-country human rights violations—we provide a simplified version of their framework in @fig-agency-response-tree.

@@

Donor countries have a range of possible responses to recipient repression, including doing nothing and ignoring the violation, threatening to change aid allocations, withdrawing or suspending aid, or shifting aid to other projects or programs. In turn, recipient governments can then respond by doing nothing and ignoring the threat or suspension, changing their behavior and ceasing the repression to improve their relationship with the donor country, or pressuring the donor country to back down and withdraw their complaint. In this simplified decision tree, there are 12 possible combinations of outcomes, and each pathway is conditional on a host of donor-country and recipient-country characteristics and contexts, including geopolitical dynamics in the recipient country's region, domestic political institutions within the donor country, public opinion in the donor country, coordination between coalitions of donors, the type of aid, and the type of human rights violation, among others [@CheesemanSwedlundOBrien-Udry:2024]. Such complexity makes it difficult to find consistent treatment effects in research that explores the overall effect of repression on aid—that is, we cannot definitively say that human rights abuses cause an overall reduction or increase in aid. Instead, newer research looks at donor responses in more specific circumstances, exploring the factors that lead to specific response pathways [@Corwin:2023].

Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political oppression. The effects of more subtle administrative crackdown on aid, including the passage of anti-NGO laws and the constricting of civic space, are less explored. Civil society plays a central role in the distribution and implementation of international aid. Governmental agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries. Outsourcing the provision of aid to NGOs allows donors to avoid aid capture in weak institutional environments and can lead to more effective program implementation [@Dietrich:2013; @CruzLabonneWright:2024]. Given the importance of NGOs in aid delivery, we have reason to expect that donor responses to civil society crackdown abroad are treated differently than more violent forms of repression, following different pathways in @fig-agency-response-tree.

@@

Though most anti-NGO laws have targeted foreign funding, others impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation [@Chaudhry:2022]. Donor countries respond differently to these types of oppression. @ChaudhryHeiss:2024 find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law [see also @ChaudhryHeiss:2022a]. @RightSpringmanWibbels:2024 find similar and more nuanced trends in donor responses: donors committed to political advocacy (i.e. those that fund democracy and civil society promotion activities) reduce funding for advocacy programs by more than 70% in response to new restrictions, and the reduction in aid persists for several years.

@@

# Private donor responses

@@

While private donations have become an important source of funding in the age of closing civic space and increased restrictions on official foreign funding, they are unlikely to full supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid. Testimony at a 2017 U.S. Senate Committee on Foreign Relations hearing maintained that "[n]either remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance" [@Runde:2017]. Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space.

@@

A growing number of organizations in the Global South have turned to community philanthropy. Community foundations facilitate democratic and transparent decision-making by allowing the local community to decide who recieves grants. This approach allows local groups to respond and be held accountable to local priorities, overcoming many criticisms of international donor aid [@Murad:2014].

@@

Foreign donors already face many criticisms from grantees, including implementing models without taking local contexts into account and prioritizing projects that are more easily quantifiable and regime compatible. In an environment of closing civic space, NGOs—especially those receiving grants from foreign donors—face even bigger risks. In a 2011 survey of 1,473 NGO respondents that had received foreign funds, 37% felt the greatest risk of receiving democracy assistance was being labeled an "agent" or "stooge" of outside forces.  [@Barkan:2012]. In a 2016 survey of over 1,000 activists from Colombia, Egypt, India, Kenya, Russia, Sudan, Turkey, Uganda, Ukraine, and Venezuela, respondents reported that their foreign donors created security and safety concerns by neglecting to take adequate measures to protect their partners, even when they have been targeted or arrested [@Miller-Dawkins:2017].

@@

Recent research shows that NGOs consider government rhetoric when making operational decisions. A survey experiment of NGOs across Cambodia, Uganda, and Serbia, found that negative NGO rhetoric significantly reduces NGO leaders' willingness to work in a community and curtails NGOs' preferences for partnering with local government authorities [@HollerbauerNagawaRobertson:2024]. Thus, NGOs not only seek to avoid working in communities repressed by the government, but also avoid working in areas where authorities use anti-NGO rhetoric. While donor responses have largely focused on navigating and adapting to repression, the use of such rhetoric raises broader questions about research and policy on the nonprofit sector: how is such rhetoric changing public attitudes and engagement with these groups? Given that anti-NGO repression and rhetoric can exacerbate burnout rates [@JoscelyneKnuckeySatterthwaite:2015], what are organizations doing to support human rights advocates, and what kinds of interventions can help support activists and professionals? In response to these concerns, organizations such as Defend Defenders, Frontline Defenders, Peace Brigades International, Civil Rights Defenders, and Protection International provide physical security, legal and digital support, and mental health resources to activists at risk. Regional initiatives such as the Mesoamerican Women Human Rights Defenders Initiative, Forum Asia, and the Euro-Mediterranean Foundation of Support for Human Rights Defenders also help provide formal protection mechanisms for activists.

@@

While examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed. To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, diplomatic actors, and local populations to address larger structural issues that exacerbate government crackdowns on civil society. Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.

>>>>

Beginning in 2003, Africa and Middle East Refugee Assistance (AMERA) International provided pro bono legal support for refugees entering Egypt. Based in London, this international non-governmental organization (INGO) maintained a small staff of legal experts in Cairo and raised the bulk of its funding from organizations in London, receiving most of its revenue from a private charitable foundation, Comic Relief UK. Despite Egyptian laws restricting INGO activities, AMERA enjoyed a positive relationship with government authorities. In the wake of the 2011 Arab Uprisings, though, the Egyptian government began strictly enforcing laws prohibiting foreign funding for civil society, blocking AMERA's access to Comic Relief donations. In 2014, AMERA shuttered its Egyptian office and redirected its programming to local projects in the United Kingdom [@Heiss:2019a].

The experiences of non-governmental organizations (NGOs) like AMERA are not uncommon. While a large literature in law and political science focuses on the role of international law, domestic legal institutions, and legal mobilization that may reduce state crackdown [@Klug:2005; @PowellStaton:2009; @Simmons:2009; @ConradRitter:2013], the last two decades have shown that states also use laws not to expand rights, but to further restrict rights and repress organizations [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020; @GlasiusSchalkDeLange:2020; @BromleySchoferLonghofer:2020; @Chaudhry:2022]. Administrative crackdown—or the use of laws to create barriers to entry, funding, and advocacy by civil society groups—has proliferated across the globe over the past few decades—both across democracies and autocracies [@Chaudhry:2022]. As of 2023, 87% of the world's population now lives in countries where the space available to civil society is closed, repressed, or obstructed [@CIVICUSMonitor].

The growth in states using law to repress can be attributed to three main factors. First, administrative or legal crackdowns have many advantages over the use of violence. It allows governments to maintain a democratic facade and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022]. Anti-NGO laws rarely elicit international condemnation or threats to withdraw aid. States also learn about the efficacy of these legal strategies from their peers and subsequently implement them in their own country [@GlasiusSchalkDeLange:2020; @Gilbert:2020; @Chaudhry:2022]. Finally, these tactics are part of a broader trend of using legal means to subvert democratic institutions and norms [@LevitskyZiblatt:2018].

Administrative crackdown via anti-NGO laws that bar access to foreign funding and foreign technical assistance has been particularly insidious. Without access to foreign funds, many INGOs, such as AMERA, cannot continue their overseas operations. Even domestic NGOs may be heavily strapped for money. Most NGOs in the Global South require a constant stream of competitive grants from foreign donors to implement specific projects [@Bush:2015]. Local groups tackling contentious issues may not be able to raise funds domestically, as publics may prefer donating to service organizations focusing on health, education, and poverty [@DupuyRonPrakash:2016; @Brechenmacher:2017]. Even local philanthropists may be deterred from giving to organizations focusing on contentious issues such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution.

@@

# Responses from official aid donors

A rich literature on the purposes of foreign aid explores why states transfer money and resources to other countries. Both national aid agencies and multilateral international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with. States rhetorically declare normative justifications for engaging in development, with recipient-focused goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011]. In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy-oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017]. A key enforcement mechanism to ensure that recipient-country actions align with donor-country goals is to condition aid on the achievement of specific democratizing and economic goals [@CheesemanSwedlundOBrien-Udry:2024; @HoefflerJustino:2024]. However, this dual-purpose strategy often creates dilemmas—if aid is withdrawn, it may jeopardize humanitarian programs or worsen political conditions. Conversely, continuing aid might signal tolerance for rights violations, weakening the credibility of the donor's human rights agenda. Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that donor agencies punish repressive states by reducing or withdrawing aid [@LebovicVoeten:2009], increasing aid to repressive states as a way to counter state violence [@Nielsen:2013], or making no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b].

One reason for such mixed results is that the decision to withdraw or suspend aid in response to human rights abuses (including civil society crackdown) is often seen as a binary choice driven by self-interested motivations: a donor country that sees human rights abuses in a recipient state will consider its own foreign policy interests before deciding to either continue or curtail aid. In reality, however, the decision is far more complex and involves a wide range of possible motivations and responses. @CheesemanSwedlundOBrien-Udry:2024 offer a stylized decision tree of possible choices following recipient-country human rights violations—we provide a simplified version of their framework in @fig-agency-response-tree.

@@

Donor countries thus have a range of possible responses to recipient repression, including doing nothing and ignoring the violation, threatening to change aid allocations, withdrawing or suspending aid, or shifting aid to other projects or programs. In turn, recipient governments can then respond by doing nothing and ignoring the suspension, or changing their behavior and ceasing the repression to improve their relationship with the donor country, or pressuring the donor country to back down and withdraw their complaint. In this simplified decision tree, there are 12 possible combinations of outcomes, and each pathway is conditional on a host of donor-country and recipient-country characteristics and contexts, including geopolitical dynamics in the recipient country's region, domestic political institutions within the donor country, public opinion in the donor country, coordination between coalitions of donors, the type of aid, and the type of human rights violation, among others [@CheesemanSwedlundOBrien-Udry:2024]. Such complexity makes it difficult to find consistent treatment effects in research that explores the overall effect of repression on aid—that is, we cannot definitively say that human rights abuses cause an overall reduction or increase in aid. Instead, newer research looks at donor responses in more specific circumstances, exploring the factors that lead to specific response pathways [@Corwin:2023].

Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political repression. The effects of more subtle administrative crackdown on aid, including the passage of anti-NGO laws and the constricting of civic space, are less explored. Civil society plays a central role in the distribution and implementation of international aid. Government agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries. Outsourcing the provision of aid to NGOs allows donors to avoid aid capture in weak institutional environments and can lead to more effective program implementation [@Dietrich:2013; @CruzLabonneWright:2024]. Given the importance of NGOs in aid delivery, we have reason to expect that donor responses to civil society crackdown abroad are treated differently than more violent forms of repression, following different pathways in @fig-agency-response-tree.

@@

Though most anti-NGO laws have targeted foreign funding, others impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation, specifically through vaguely defined "political activities" [@Chaudhry:2022]. Donor countries respond differently to these types of oppression. @ChaudhryHeiss:2024 find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law [see also @ChaudhryHeiss:2022a]. @RightSpringmanWibbels:2024 find similar and more nuanced trends in donor responses: donors committed to political advocacy (i.e. those that fund democracy and civil society promotion activities) reduce funding for advocacy programs by more than 70% in response to new restrictions, and the reduction in aid persists for several years.

@@

# Responses from private donors

@@

While private donations have become an important source of funding in the age of closing civic space and increased restrictions on foreign funding, they are unlikely to fully supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid. Testimony at a 2017 U.S. Senate Committee on Foreign Relations hearing maintained that "[n]either remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance" [@Runde:2017]. Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space.

@@

A growing number of organizations in the Global South have also turned to community philanthropy. Community foundations facilitate democratic and transparent decision-making by allowing the local community to decide who recieves grants. This approach allows local groups to respond and be held accountable to local priorities, overcoming many criticisms of international donor aid [@Murad:2014].

@@

Foreign donors already face many criticisms from grantees, including implementing models without taking local contexts into account and prioritizing projects that are more easily quantifiable and regime compatible [@Bush:2015]. In an environment of closing civic space, NGOs—especially those receiving grants from foreign donors—face even bigger risks. In a 2011 survey of 1,473 NGO respondents that had received foreign funds, 37% felt the greatest risk of receiving democracy assistance was being labeled an "agent" or "stooge" of outside forces.  [@Barkan:2012]. In a 2016 survey of over 1,000 activists from Colombia, Egypt, India, Kenya, Russia, Sudan, Turkey, Uganda, Ukraine, and Venezuela, respondents reported that their foreign donors created security and safety concerns by neglecting to take adequate measures to protect their partners, even when they have been targeted or arrested [@Miller-Dawkins:2017].

@@

Recent research shows that NGOs consider government rhetoric when making operational decisions. A survey experiment of NGOs across Cambodia, Uganda, and Serbia, found that negative NGO rhetoric significantly reduces NGO leaders' willingness to work in a community and curtails NGOs' preferences for partnering with local government authorities [@HollerbauerNagawaRobertson:2024]. Thus, NGOs not only seek to avoid working in communities repressed by the government, but also avoid working in areas where authorities use anti-NGO rhetoric. While donor responses have largely focused on navigating and adapting to repression, the use of such rhetoric raises broader questions about research and policy in the nonprofit sector: how is such rhetoric changing public attitudes and engagement with these groups? Given that anti-NGO repression and rhetoric can exacerbate burnout rates [@JoscelyneKnuckeySatterthwaite:2015], what are organizations doing to support rights advocates, and what kinds of interventions can help support activists and professionals? In response to these concerns, organizations such as Defend Defenders, Frontline Defenders, Peace Brigades International, Civil Rights Defenders, and Protection International provide physical security, legal and digital support, and mental health resources to activists at risk. Regional initiatives such as the Mesoamerican Women Human Rights Defenders Initiative, Forum Asia, and the Euro-Mediterranean Foundation of Support for Human Rights Defenders also help provide formal protection mechanisms for activists.

@@

While the examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed. To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, and local populations to address larger structural issues that exacerbate government crackdowns on civil society. Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.
